GMSE: an R package for generalised management strategy evaluation

### 2017-08-08 notebook/ms/ms.Rmd
Major restructuring of Utility of GMSE section moved to the Introduction

The GMSE R package is a flexible, highly mechanistic, agent-based modelling tool to simulate all key aspects of natural resource management. GMSE considers a range of parameters to simulate resource dynamics and management policy options, and uses genetic algorithms to dynamically model stakeholder (manager and user) decision-making. Genetic algorithms allow stakeholders to find adaptive solutions to any simulated conditions given unique agent goals, allowing GMSE to model scenarios of conservation conflict.

@@

Here we consider the example of a managed natural resource whose abundance affects a group of stakeholders by temporarily decreasing the value of user land. This scenario could be interpreted in multiple ways; we consider a protected population of waterfowl that exploits agricultural land causing a conservation conflict with users [hereafter 'farmers', e.g., @Tuvendal2015; @Fox2016a; @Fox2017a]. Managers in this example might want to keep the abundance of waterfowl at a target level, while farmers might want to minimise the damage inflicted on their crops. Using GMSE, it is possible to simulate waterfowl population dynamics, along with the continued monitoring and policy set by managers, and the actions that farmers take to protect their crop yields given the constraints of policy. Here we consider a population of waterfowl with an initial abundance and manager target abundance of 1000, but whose carrying capacity (applied to mortality) is 2000. Waterfowl consume and destroy all crop yield upon arrival to a landscape cell. Managers estimate population size in each time step using mark-recapture techniques, and use their estimates to set costs of culling and scaring (non-lethal) waterfowl (i.e., 'policy') for ten farmers. Farmers attempt to reduce the negative impact of waterfowl on the cropland that they own, working within the constraints of culling and scaring costs and their budget for performing these actions.

@@

Only the estimates of population size from the observation model are available to the manager, so policy change at any time step is driven primarily by the deviation of the currently estimated population size from the manager's target and the actions of farmers in the previous time step. Hence, when the population size is estimated to be below (above) the manager's target, the manager increases (decreases) the cost of culling and decreases (increases) the cost of scaring. Because the manager does not know in advance how stakeholders will react to policy change, they assume a proportional response in total actions with respect to a change in cost (e.g., doubling the cost of culling will decrease stakeholder culling by $1/2$). Farmers responding to policy are interested only in mitigating waterfowl's exploitation of their crops, so they will either cull or scare to remove the waterfowl from their land, depending on which option is more effective (i.e., cheaper). This is reflected in the bottom left versus right panels of Figure 2; when managers decrease culling costs relative to scaring, farmers respond with more total culling, and vice versa. Farmer decisions then affect waterfowl distribution and abundance, impacting future crop yield and policy.

Utility of GMSE simulations
===================================================

GMSE incorporates feedback between resource dynamcs, management policy, and user actions over time given multiple sources of uncertainty. By simulating complex systems of adaptive management in GMSE, it is possible to gain new insights into questions relevant to conservation. One obvious application of our software is to parameterise simulations with initial conditions similar to those from empirical populations of conservation interest. Running replicate simulations using empirically derived starting conditions could facilitate the prediction of key social-ecological outcomes (e.g., resource extinction, agricultural yield) given uncertainty, and the sensitivity of such outcomes to different management options (e.g., population target, policies available, observation methods, etc.), realised through results from replicate simulations, could inform management decisions.

Additionally, GMSE might be useful for developing adaptive management theory by allowing researchers to explore general questions concerning management *in silico* through simulation. Such questions might include the following: How is population persistence affected by management frequency or observation intensity? How does variation in user actions affect the distribution of resources or landscape properties? How do asymmetries in stakeholder influence (i.e., budgets) affect resource dynamics? All of these questions can be explored through simulation using GMSE.

>>>>

The GMSE R package is a flexible, highly mechanistic, agent-based modelling tool to simulate all key aspects of natural resource management. GMSE considers a range of parameters to simulate resource dynamics and management policy options, and uses genetic algorithms to dynamically model stakeholder (manager and user) decision-making. Genetic algorithms allow stakeholders to find adaptive solutions to any simulated conditions given unique agent goals, allowing GMSE to model scenarios of conservation conflict. Hence, simulations can be parameterised with initial conditions similar to those from empirical populations of conservation interest to predict key social-ecological outcomes (e.g., resource extinction, agricultural yield) given uncertainty. The sensitivity of these outcomes to different management options (e.g., population target, policies available, observation methods, budget constraints, etc.) can thereby inform decisions underpinning competing management objectives.  Additionally, GMSE allows researchers to explore general questions concerning management *in silico* through simulation. Such questions might include the following: How is population persistence affected by management frequency or observation intensity? How does variation in user actions affect the distribution of resources or landscape properties? How do asymmetries in stakeholder influence (i.e., budgets) affect resource dynamics?

@@

Here we consider the example of a managed natural resource whose abundance affects a group of stakeholders by temporarily decreasing the value of user land. This scenario could be interpreted in multiple ways; we consider a protected population of waterfowl that exploits agricultural land causing a conservation conflict with users [hereafter 'farmers', e.g., @Tuvendal2015; @Fox2016a; @Fox2017a]. Managers in this example might want to keep the abundance of waterfowl at a target level, while farmers might want to minimise the damage inflicted on their crops [e.g., @Madsen2017]. Using GMSE, it is possible to simulate waterfowl population dynamics, along with the continued monitoring and policy set by managers, and the actions that farmers take to protect their crop yields given the constraints of policy. Here we consider a population of waterfowl with an initial abundance and manager target abundance of 1000, but whose carrying capacity is 2000. Waterfowl consume and destroy all crop yield upon arrival to a landscape cell. In each time step, waterfowl are observed on subset of cells, then managers extrapolate from density per cell to estimate total population size. Managers then use these estimates to set costs of culling and scaring (non-lethal) waterfowl (i.e., 'policy') for 10 farmers. Farmers attempt to reduce the negative impact of waterfowl on the cropland that they own, working within the constraints of culling and scaring costs and their budget for performing these actions.

@@

Only the estimates of population size from the observation model are available to the manager, so policy change at any time step is driven primarily by the deviation of the currently estimated population size from the manager's target and the actions of farmers in the previous time step. Hence, when the population size is estimated to be below (above) the manager's target, the manager increases (decreases) the cost of culling and decreases (increases) the cost of scaring. Because the manager does not know in advance how stakeholders will react to policy change, they assume a proportional response in total actions with respect to a change in cost (e.g., doubling the cost of culling will decrease stakeholder culling by $1/2$). Farmers responding to policy are interested only in minimising waterfowl's exploitation of their crops, so they will either cull or scare to remove the waterfowl from their land, depending on which option is more effective (i.e., cheaper). This is reflected in the bottom left versus right panels of Figure 2; when managers decrease culling costs relative to scaring, farmers respond with more total culling, and vice versa. Farmer decisions then affect waterfowl distribution and abundance, impacting future crop yield and policy.
